
# Installed required packages
```{r}
install.packages("matlib")
install.packages("rsample")
```

# Imported libraries
```{r}
library(matlib)
library(ggplot2)
library(rsample)
```

# Reading the dataset
```{r}
data <- read.csv("/Users/kriipa/Downloads/dataset.csv")
str(data)
head(data)
```


# Checking for missing values in each column
```{r}

null_Values <- data.frame(
  Missing_Values = colSums(is.na(data))
)

print(null_Values)

```

# Check if any duplicates
```{r}
# Count duplicate rows
sum(duplicated(data))

```

# Display a summary of the dataset, showing basic statistics for numerical variables and frequency counts for categorical variables.
```{r}
summary(data)
```


# Separated the dataset into independent variables (X) and dependent variable (Y).
# Displayed the first few rows of both X and Y to check the separation.
```{r}
X <- data[, c("x1", "x3", "x4", "x5")]  # Independent variables
Y <- data[, "x2", drop = FALSE]         # Dependent variable

# Check
head(X)
head(Y)

```

#2.1

```{r}

# Created an observation index and added it as a new column to the dataset.
data$Observation <- 1:nrow(data)

# Subsetted the first 500 observations for analysis.
data_sub <- data[1:500, ]

# Plotted time series for each variable (x1, x3, x4, x5, x2) using ggplot2:
# Each plot includes appropriate labels and color schemes to visualize trends for each variable.

# Input: x1 - Temperature
ggplot(data_sub,   aes(x = Observation, y = x1)) +
  geom_line( color="navy") +
  labs(title = "Time Series: Ambient Temperature (x1)", 
       x = "Observation", y = "Temperature (°C)")

# Input: x3 - Ambient Pressure
ggplot(data_sub, aes(x = Observation, y = x3)) +
  geom_line( color = "maroon") +
  labs(title = "Time Series: Ambient Pressure (x3)",
       x = "Observation", y = "Pressure (millibar)")

# Input: x4 - Relative Humidity
ggplot(data_sub, aes(x = Observation, y = x4)) +
  geom_line( color = "forestgreen") +
  labs(title = "Time Series: Relative Humidity (x4)",
       x = "Observation", y = "Humidity (%)")

# Input: x5 - Exhaust Vacuum
ggplot(data_sub, aes(x = Observation, y = x5)) +
  geom_line( color = "gold") +
  labs(title = "Time Series: Exhaust Vacuum (x5)",
       x = "Observation", y = "Vacuum (cm Hg)")

# Output: x2 - Net hourly electrical energy output
ggplot(data_sub, aes(x = Observation, y = x2)) +
  geom_line( color = "brown") +
  labs(title = "Time Series: Net Hourly Electrical Energy Output (x2)",
       x = "Observation", y = "Energy Output (MW)")


```

# Generated histograms and density plots for each variable (x1, x3, x4, x5, x2).
# For each variable, a histogram with density scaling was created, overlaid with a density plot to visualize the distribution.

```{r}
 # A map  to associate each variable with its physical meaning for title labeling.
for (col in c("x1", "x3", "x4", "x5", "x2")) {
  var_map <- c(
    x1 = "Ambient Temperature (°C)", 
    x3 = "Ambient Pressure (millibar)", 
    x4 = "Relative Humidity (%)", 
    x5 = "Exhaust Vacuum (cm Hg)", 
    x2 = "Net Hourly Electrical Energy Output (MW)"
  )
  
# For each variable, a histogram with 30 bins and a density plot were displayed to visualize the distribution and spread of data.
  print(
    ggplot(data, aes_string(x = col)) +
      geom_histogram(aes(y = ..density..), bins = 30, fill = "peachpuff3", color = "black", alpha = 0.7) +
      geom_density(color = "maroon", linewidth = 1) +
      labs(
        title = paste("Histogram and Density of", var_map[col]),
        x = var_map[col],
        y = "Density"
      )
  )
}

```


# 1.2

```{r}
# Calculate and round correlation matrix
cor_matrix <- cor(data[, c("x1", "x3", "x4", "x5", "x2")])

# The matrix was rounded to two decimal places for clarity.
cor_table <- round(cor_matrix, 2)

# Converted the correlation matrix to a data frame for a more readable table output.
cor_df <- as.data.frame(cor_table)

print(cor_df)
```

# Plotted scatter plots for each input variable against the dependent variable (x2).
# Each scatter plot included a linear regression line to show the relationship between the input and output variables.
```{r}
inputs <- c("x1", "x3", "x4", "x5")
var_map <- c(
  x1 = "Ambient Temperature (°C)", 
  x3 = "Ambient Pressure (millibar)", 
  x4 = "Relative Humidity (%)", 
  x5 = "Exhaust Vacuum (cm Hg)", 
  x2 = "Net Hourly Electrical Energy Output (MW)"
)

for (input in inputs) {
  print(
    ggplot(data, aes_string(x = input, y = "x2")) +
      geom_point(alpha = 0.5, color="peachpuff4") +
      geom_smooth(method = "lm", color = "maroon", se = FALSE, linewidth=1) +
      labs(
        title = paste(var_map["x2"], "vs", var_map[input]),
        x = var_map[input],
        y = var_map["x2"]
      )
  )
}



```

# Correlation matrix
```{r}
# Create a correlation matrix dataframe for ggplot2
inputs_outputs <- c("x1","x3","x4","x5","x2")

cor_matrix <- cor(data[, inputs_outputs])
cor_df <- as.data.frame(as.table(cor_matrix))

ggplot(cor_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), color = "white", size = 4) +
  labs(
    title = "Correlation Heatmap",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "pink", midpoint = 0)

```

# Visualized the relationship between temperature (x1) and energy output (x2) with scatter points.
```{r}
ggplot(data, aes(x = x1, y = x2)) +
  geom_point(alpha = 0.4, color="seagreen") +
  geom_density2d(color = "black") +
  labs(title = "Density Contours: Temperature vs Energy Output",
       x = "Temperature (°C)", y = "Energy Output (MW)")

```

# Explored how humidity (x4) influences the relationship between temperature (x1) and energy output (x2).
# Created a scatter plot with points colored based on humidity levels to visualize this effect.
```{r}
# analyzing how humidity influences the temperature-energy output relationship:
ggplot(data, aes(x = x1, y = x2, color = x4)) +
  geom_point(alpha = 0.5) +
  labs(title = "Temperature vs Energy Output (Colored by Humidity)",
       x = "Temperature (°C)", y = "Energy Output (MW)", color = "Humidity (%)")

```


## TASK 2.1

# Defined a helper function to compute least squares coefficients using generalized inverse.
```{r}
least_squares_coef <- function(x, y) {
  MASS::ginv(t(x) %*% x) %*% t(x) %*% y
}

Y <- data[, "x2"]

# Model fits
model1 <- as.matrix(cbind(x4 = data$x4, x3_sq = data$x3^2, bias = 1))
theta_1 <- as.numeric(least_squares_coef(model1, Y))

model2 <- as.matrix(cbind(x4 = data$x4, x3_sq = data$x3^2, x5 = data$x5, bias = 1))
theta_2 <- as.numeric(least_squares_coef(model2, Y))

model3 <- as.matrix(cbind(x3 = data$x3, x4 = data$x4, x5_cub = data$x5^3))
theta_3 <- as.numeric(least_squares_coef(model3, Y))

model4 <- as.matrix(cbind(x4 = data$x4, x3_sq = data$x3^2, x5_cub = data$x5^3, bias = 1))
theta_4 <- as.numeric(least_squares_coef(model4, Y))

model5 <- as.matrix(cbind(x4 = data$x4, x1_sq = data$x1^2, x3_sq = data$x3^2, bias = 1))
theta_5 <- as.numeric(least_squares_coef(model5, Y))

# Build a table
coef_table <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  `θ1` = c(theta_1[1], theta_2[1], theta_3[1], theta_4[1], theta_5[1]),
  `θ2` = c(theta_1[2], theta_2[2], theta_3[2], theta_4[2], theta_5[2]),
  `θ3` = c(NA, theta_2[3], theta_3[3], theta_4[3], theta_5[3]),
  Bias = c(theta_1[3], theta_2[4], NA, theta_4[4], theta_5[4])
)

# Round for clarity
coef_table[, -1] <- round(coef_table[, -1], 4)

# printed
cat("Task 2.1: Model Coefficient Estimates Table\n")
print(coef_table, row.names = FALSE, na.print = "")

```


# Task 2.2

# Computed the Residual Sum of Squares (RSS) for each model to measure the fit error.
```{r}
# Task 2.2: Compute RSS for all models

# Model 1
y_hat_1 <- model1 %*% theta_1
rss_1 <- sum((Y - y_hat_1)^2)

# Model 2
y_hat_2 <- model2 %*% theta_2
rss_2 <- sum((Y - y_hat_2)^2)

# Model 3 (no bias)
y_hat_3 <- model3 %*% theta_3
rss_3 <- sum((Y - y_hat_3)^2)

# Model 4
y_hat_4 <- model4 %*% theta_4
rss_4 <- sum((Y - y_hat_4)^2)

# Model 5
y_hat_5 <- model5 %*% theta_5
rss_5 <- sum((Y - y_hat_5)^2)

# Make a neat table
rss_table <- data.frame(
  Model = paste0("Model ", 1:5),
  RSS = round(c(rss_1, rss_2, rss_3, rss_4, rss_5), 4)
)

cat("Task 2.2: Residual Sum of Squares (RSS) Table\n")
print(rss_table, row.names = FALSE)
```


# Task 2.3
# Defined a function to calculate the log-likelihood based on RSS and sample size.
```{r}
n <- nrow(data)
loglik_fun <- function(RSS, n) {
  sigma2_hat <- RSS / (n - 1)
  term1 <- -n / 2 * log(2 * pi)
  term2 <- -n / 2 * log(sigma2_hat)
  term3 <- -RSS / (2 * sigma2_hat)
  term1 + term2 + term3
}

loglik1 <- loglik_fun(rss_1, n)
loglik2 <- loglik_fun(rss_2, n)
loglik3 <- loglik_fun(rss_3, n)
loglik4 <- loglik_fun(rss_4, n)
loglik5 <- loglik_fun(rss_5, n)

# Created a table summarizing the log-likelihoods and identified the best model with the highest value.
loglik_values <- c(loglik1, loglik2, loglik3, loglik4, loglik5)
loglik_table <- data.frame(
  Model = paste0("Model ", 1:5),
  LogLikelihood = round(loglik_values, 4)
)

# Find best model (highest log-likelihood)
best_idx <- which.max(loglik_table$LogLikelihood)
loglik_table$Best <- ""
loglik_table$Best[best_idx] <- "<-- Highest"

cat("Task 2.3: Log-Likelihood Table\n")
print(loglik_table, row.names = FALSE)

# Higher log-likelihood values (less negative) indicate a better fit to the data
```



# Task 2.4
# Calculated AIC and BIC values for all models using log-likelihood and parameter counts.
```{r}
# Number of parameters in each model
k <- c(3, 4, 3, 4, 4)

# Compute AIC and BIC for each model
aic_values <- 2 * k - 2 * loglik_values
bic_values <- k * log(n) - 2 * loglik_values

# Make the table
info_criteria_table <- data.frame(
  Model = paste0("Model ", 1:5),
  aic = round(aic_values, 4),
  best_aic = "",
  bic = round(bic_values, 4),
  best_bic = ""
)

# Indicate best (lowest) AIC and BIC
info_criteria_table$best_aic[which.min(info_criteria_table$aic)] <- "<-- Lowest"
info_criteria_table$best_bic[which.min(info_criteria_table$bic)] <- "<-- Lowest"

cat("Task 2.4: AIC and BIC Table")
print(info_criteria_table, row.names = FALSE)

```


# Task 2.5
# Plotted histograms and Q-Q plots of residuals for all models to assess normality and distribution.
```{r}
residuals1 <- Y - as.vector(model1 %*% theta_1)
residuals2 <- Y - as.vector(model2 %*% theta_2)
residuals3 <- Y - as.vector(model3 %*% theta_3)
residuals4 <- Y - as.vector(model4 %*% theta_4)
residuals5 <- Y - as.vector(model5 %*% theta_5)

# Plot histograms and Q-Q plots in one window
par(mfrow = c(1, 1), mar = c(4, 4, 2, 1)) # 5 rows, 2 columns

# Model 1
hist(residuals1, main = "Model 1 Residuals Histogram", xlab = "Residuals", col="seagreen")
qqnorm(residuals1, main = "Model 1 Q-Q Plot", col="peachpuff4")
qqline(residuals1)
# Model 2
hist(residuals2, main = "Model 2 Residuals Histogram", xlab = "Residuals", col="seagreen")
qqnorm(residuals2, main = "Model 2 Q-Q Plot", col="peachpuff4")
qqline(residuals2)
# Model 3
hist(residuals3, main = "Model 3 Residuals Histogram", xlab = "Residuals", col="seagreen")
qqnorm(residuals3, main = "Model 3 Q-Q Plot", col="peachpuff4")
qqline(residuals3)
# Model 4
hist(residuals4, main = "Model 4 Residuals Histogram", xlab = "Residuals", col="seagreen")
qqnorm(residuals4, main = "Model 4 Q-Q Plot", col="peachpuff4")
qqline(residuals4)
# Model 5
hist(residuals5, main = "Model 5 Residuals Histogram", xlab = "Residuals", col="seagreen")
qqnorm(residuals5, main = "Model 5 Q-Q Plot", col="peachpuff4")
qqline(residuals5)

par(mfrow = c(1, 1)) # Reset layout
```


# TASK 2.6
# Sampled up to 5000 residuals from each model if residuals exceeded that size.

```{r}
set.seed(42) # for reproducibility
sample_size <- 5000

# Only sample if your residual vector is larger than 5000
sample1 <- if (length(residuals1) > sample_size) sample(residuals1, sample_size) else residuals1
sample2 <- if (length(residuals2) > sample_size) sample(residuals2, sample_size) else residuals2
sample3 <- if (length(residuals3) > sample_size) sample(residuals3, sample_size) else residuals3
sample4 <- if (length(residuals4) > sample_size) sample(residuals4, sample_size) else residuals4
sample5 <- if (length(residuals5) > sample_size) sample(residuals5, sample_size) else residuals5

shapiro1 <- shapiro.test(sample1)
shapiro2 <- shapiro.test(sample2)
shapiro3 <- shapiro.test(sample3)
shapiro4 <- shapiro.test(sample4)
shapiro5 <- shapiro.test(sample5)

cat("Shapiro-Wilk p-values:\n")
cat("Model 1:", shapiro1$p.value, "\n")
cat("Model 2:", shapiro2$p.value, "\n")
cat("Model 3:", shapiro3$p.value, "\n")
cat("Model 4:", shapiro4$p.value, "\n")
cat("Model 5:", shapiro5$p.value, "\n")

```

# 2.7
# Split the dataset into training (70%) and testing (30%) subsets using random sampling.
#1
```{r}
# 1. Split Data (70% train, 30% test)
set.seed(123)
n <- nrow(data)
train_idx <- sample(seq_len(n), size = 0.7 * n)

# Training and testing splits
train_data <- data[train_idx, ]
test_data  <- data[-train_idx, ]
```

# Prepared the training input matrix by combining variables x4, squared terms of x1 and x3, and a bias term.
#2
```{r}
# Training:
X5_train <- as.matrix(cbind(train_data$x4, train_data$x1^2, train_data$x3^2, 1))
Y_train <- train_data$x2
```


# Testing
```{r}
X5_test <- as.matrix(cbind(test_data$x4, test_data$x1^2, test_data$x3^2, 1))
Y_test <- test_data$x2
```

#3. Fit the Model (Estimate Parameters)
```{r}
#This is an ordinary least squares regression:
theta_hat5 <- solve(t(X5_train) %*% X5_train) %*% t(X5_train) %*% Y_train
```


# 4. Compute 95% Prediction Intervals
# Predicted output values on the test set using the trained model coefficients.
```{r}
#Predict on Test Data

Y_pred <- as.vector(X5_test %*% theta_hat5)

# Here’s a method that works for general linear models (not using lm, but your matrix approach):

# Calculate sigma^2 (estimate of variance)
residuals_train <- Y_train - as.vector(X5_train %*% theta_hat5)
sigma2_hat <- sum(residuals_train^2) / (nrow(X5_train) - ncol(X5_train))

# Standard errors for predictions
XtX_inv <- solve(t(X5_train) %*% X5_train)
se_pred <- sqrt(apply(X5_test, 1, function(xi) {
    sigma2_hat * (1 + t(xi) %*% XtX_inv %*% xi)
}))

# 95% interval (z-value ≈ 1.96)
lower <- Y_pred - 1.96 * se_pred
upper <- Y_pred + 1.96 * se_pred
```

# 5. Plot Results
```{r}
# Setting transluscent color for arrows
transparent_col <- adjustcolor("peachpuff4", alpha.f = 0.6)

# Y_test is test data
plot(Y_test, pch = 16, col = "black", xlab = "Test Sample", ylab = "x2", main = "Model 5 Predictions with 95% Interval")

# Y_pred is predicted data
points(Y_pred, pch = 16, col = "plum")

# arrows indicates 95% interval
arrows(1:length(Y_pred), lower, 1:length(Y_pred), upper, angle = 90, code = 3, length = 0.03, col = transparent_col)

# indexing for clarity
legend("topright", legend = c("Observed", "Predicted", "95% Interval"), col = c("black", "plum", "peachpuff4"), pch = c(16, 16, NA), lty = c(NA, NA, 1))

```

# Calculated the width of the 95% prediction intervals for the test set.
```{r}
interval_width <- upper - lower
plot(interval_width,
    main = "Width of 95% Prediction Intervals (Test Set)",
    xlab = "Test Sample Index", ylab = "Interval Width",
    pch = 16, col = "plum"
)
```

#------- task 3-----


# 1. Identify 2 Parameters with Largest Absolute Value
```{r}
abs_theta <- abs(theta_5)
largest_idx <- order(abs_theta, decreasing = TRUE)[1:2]
cat("Indices of largest coefficients:", largest_idx, "\n")
cat("Values:", theta_5[largest_idx], "\n")
```


# 2. Set Uniform Priors Around Their Estimates
```{r}
prior_range <- 0.2 # +/- 20%
param1_hat <- theta_5[largest_idx[1]]
param2_hat <- theta_5[largest_idx[2]]

param1_prior <- c(param1_hat * (1 - prior_range), param1_hat * (1 + prior_range))
param2_prior <- c(param2_hat * (1 - prior_range), param2_hat * (1 + prior_range))
```

# Fix All Other Coefficients to Least Squares Estimates
```{r}
fixed_thetas <- theta_5
fixed_thetas[largest_idx] <- NA # these two vary in ABC

X_abc <- model5 # your design matrix for Model 5
Y_abc <- as.vector(Y) # true outputs
```

# Run Rejection ABC
```{r}
set.seed(123)
N <- 10000

# Ensure param priors are sorted properly
param1_prior <- sort(param1_prior)
param2_prior <- sort(param2_prior)

# Sample from uniform priors
param1_samples <- runif(N, min = param1_prior[1], max = param1_prior[2])
param2_samples <- runif(N, min = param2_prior[1], max = param2_prior[2])

distances <- numeric(N)

# Check for NA in input data once
if (anyNA(X_abc)) stop("X_abc contains NA values!")
if (anyNA(Y_abc)) stop("Y_abc contains NA values!")

for (i in 1:N) {
  theta <- fixed_thetas
  
  # Replace NA with 0 for safe multiplication
  theta[is.na(theta)] <- 0
  
  # Update theta with current samples
  theta[largest_idx[1]] <- param1_samples[i]
  theta[largest_idx[2]] <- param2_samples[i]
  
  Y_sim <- as.vector(X_abc %*% theta)
  
  # Guard against NA or NaN in prediction
  if (any(is.na(Y_sim)) || any(is.nan(Y_sim))) {
    distances[i] <- NA
  } else {
    distances[i] <- sum((Y_sim - Y_abc)^2)
  }
}

# Filter out NA distances before quantile calculation
valid_idx <- which(!is.na(distances))
distances_valid <- distances[valid_idx]

# Calculate epsilon (1% quantile)
epsilon <- quantile(distances_valid, 0.01)

# Select accepted samples based on epsilon threshold
accepted_valid <- valid_idx[which(distances_valid <= epsilon)]

posterior_param1 <- param1_samples[accepted_valid]
posterior_param2 <- param2_samples[accepted_valid]

```

# 4. Plot Posterior Marginals and Joint Distribution
```{r}
par(mfrow = c(1, 1))
hist(posterior_param1,
     breaks = 30, col = "lightblue",
     main = paste("Marginal Posterior: Parameter", largest_idx[1]),
     xlab = "Parameter Value"
)
hist(posterior_param2,
     breaks = 30, col = "seagreen",
     main = paste("Marginal Posterior: Parameter", largest_idx[2]),
     xlab = "Parameter Value"
)

par(mfrow = c(1, 1))

plot(posterior_param1, posterior_param2,
     pch = 16, cex = 0.5,
     col = rgb(0, 0, 1, 0.3), main = "Joint Posterior Distribution",
     xlab = paste("Parameter", largest_idx[1]),
     ylab = paste("Parameter", largest_idx[2])
)
```


```{r}
library(rmarkdown)
render("KripaDatatask.Rmd", output_format = "md_document")
```

